---
title: "Whittakerr — Biome Maps in Geographic Space"
output: html_notebook
---

## Overview

This notebook demonstrates `map_biomes()` — a function that
produces a geographic biome map for a region by combining
climate retrieval (per `get_climate()`) with biome
classification (per `name_biome()`) and rendering the result
over a region polygon.

Where `plot_biomes()` plots data points in T-P diagram space,
`map_biomes()` plots biome assignments back in real
geographic space, completing the classification-to-mapping
arc that the document's design notes describe as the
project's destination.

The first worked example is Oahu — small enough to iterate
quickly, dramatic enough (windward-leeward rain shadow,
elevational gradient on the Koolau range) to actually exercise
the resolution choices.

## Setup

```{r}

## Load the whittakerr package. Provides map_biomes(),
## smooth_biome_map(), plot_biome_map(), name_biome(),
## plot_biomes(), and the bundled data (Whittaker_biomes,
## Ricklefs_colors).
##
## If whittakerr isn't installed yet:
##   library(devtools)
##   install_github("kimbridges/whittakerr")

library(whittakerr) ## the canonical source
library(geodata)    ## WorldClim and GADM download
library(terra)      ## raster ops, polygon ops, plot
library(tibble)     ## return types
library(dplyr)      ## piping, mutate
library(ggplot2)    ## plots package
library(sp)         ## point.in.polygon (used by name_biome)
library(sf)         ## simple features (smoothr interop)
library(smoothr)    ## polygon boundary smoothing

```

## Fetching the Oahu polygon

The region polygon for Oahu comes from GADM (Global
Administrative Areas) via the `geodata` package. Oahu is part
of Honolulu County, Hawaii. The county legally includes the
Northwestern Hawaiian Islands, so a bounding-box crop to the
main island is needed.

```{r}

## Download (or load from cache) US county boundaries.
## First call downloads; subsequent calls reuse.
usa_counties <- geodata::gadm(country = "USA",
                              level   = 2,
                              path    = "cache/gadm_cache")

## Filter to Honolulu County (Hawaii).
honolulu_co <- usa_counties[usa_counties$NAME_1 == "Hawaii" &
                            usa_counties$NAME_2 == "Honolulu", ]

## Crop to Oahu's main-island bounding box.
## NW Hawaiian Islands (Necker, French Frigate Shoals, etc.)
## are part of the county but should be excluded.
oahu_bbox <- terra::ext(-158.3, -157.6, 21.2, 21.8)
oahu      <- terra::crop(honolulu_co, oahu_bbox)

## Quick sanity check: plot the polygon.
plot(oahu, main = "Oahu boundary from GADM")

```

## Smoke test: Oahu at 2.5 arc-minute resolution

A first run at 2.5' to validate the pipeline. The cache from
the climate retrieval work is already populated at this
resolution, so this should run quickly. At 2.5', cells are
about 4.3 × 4.6 km — coarse for an island the size of Oahu;
expect a handful of cells with limited biome variation.

```{r}

oahu_map_25 <- map_biomes(region_polygon = oahu,
                          resolution     = 2.5)

## Inspect the result.
oahu_map_25$resolution
oahu_map_25$biome_names
terra::ncell(oahu_map_25$biome_raster)
table(terra::values(oahu_map_25$biome_raster),
      useNA = "ifany")

```

Plot the 2.5' map.

```{r}

plot_biome_map(oahu_map_25)

```

The 2.5' version is essentially a proof-of-pipeline. A few
cells, low spatial resolution, but the function works
end-to-end: climate retrieved, biome classified, rendered
over the Oahu coastline.

## The real version: Oahu at 30 arc-second resolution

At 30 arc-seconds (~1 km cells), Oahu becomes a respectable
map with several hundred cells. The 30" data is distributed
by WorldClim as 30°×30° tiles; the first call downloads the
tile covering Hawaii (a one-time cost, several hundred MB),
subsequent calls reuse the cache.

The windward-leeward rain-shadow signal that 2.5' averaged
away should be visible at 30".

```{r}

oahu_map_30s <- map_biomes(region_polygon = oahu,
                           resolution     = 0.5)

## Inspect.
oahu_map_30s$resolution
terra::ncell(oahu_map_30s$biome_raster)
table(terra::values(oahu_map_30s$biome_raster),
      useNA = "ifany")

```

```{r}

plot_biome_map(oahu_map_30s)

```

## What we expect to see

At 30" resolution on Oahu:

- **Windward (north and east coasts, Koolau range):**
  cooler, much wetter. Likely classified as tropical rain
  forest or temperate rain forest depending on elevation.
- **Leeward (south and west coasts):** warmer, much drier.
  Likely classified as subtropical desert or
  woodland/shrubland in the driest stretches.
- **Central plain (Waianae-to-Koolau saddle):** intermediate.
  Likely tropical seasonal forest/savanna.
- **Mt. Kaala and the Koolau peaks:** locally cooler;
  resolution-dependent on whether the highest pixels register.

If the assignments differ substantially from this, the
likeliest causes are: (a) the windward-leeward signal still
gets averaged within 1 km cells in places where the
transition is sharper than the cell size; (b) the
Whittaker biome polygons in T-P space don't quite cover the
edge cases Hawaii presents (Hawaii's volcanic-soil tropical
climates don't always map cleanly to the global
classification scheme); (c) WorldClim's data has its own
representativeness gaps in the Hawaiian archipelago.

## Smoothing the biome boundaries

The gridded version is faithful to the data — each cell is
visible at its own resolution, the pixelated appearance
reinforces that the map is a tessellation rather than a
continuous classification. But the Whittaker diagram itself
has smooth biome polygons, and published biome maps are
generally smooth. The visual dissonance between a
pixelated geographic map and a smooth diagram is worth
addressing if the smoothing doesn't cost us data fidelity.

Vector smoothing operates on the polygon representation
of the classified raster, not on the underlying climate
data. The subtropical desert finding at Kaena Point
survives because the per-cell biome assignments are
unchanged; only how their boundaries are drawn changes.

```{r}

oahu_map_30s_smooth <- smooth_biome_map(
  biome_map  = oahu_map_30s,
  method     = "ksmooth",
  smoothness = 1
)

```

Compare gridded and smoothed at the same 30 arc-second
resolution. Each plot gets its own chunk so the legend has
room to display the full biome names without truncation.

```{r fig.width=10, fig.height=6}

plot_biome_map(oahu_map_30s_smooth, render = "grid")

```

```{r fig.width=10, fig.height=6}

plot_biome_map(oahu_map_30s_smooth, render = "vector")

```

What to look for:

- **Aesthetic continuity** with the Whittaker diagram —
  smooth boundaries everywhere rather than pixelated
  cell edges.
- **Survival of the subtropical desert finding** — the
  Kaena Point and Waianae-lee dry pockets should still
  be visible as desert polygons, just with smooth
  boundaries rather than pixelated edges.
- **Possible artifacts** — small isolated cells may
  produce small smoothed polygons that look unnatural;
  if a biome appears as several scattered tiny blobs in
  the gridded view, the smoothed view may amplify
  the visual impact of those blobs in a misleading way.
- **The verdict** — does the smoothed version look like
  a published biome map, or does it look like
  over-processed output? That's the decision criterion
  for whether smoothing earns a place in the toolkit
  and the chapter.

If smoothing doesn't earn its place after this comparison,
the function stays in the code but defaults to off, and
the chapter discusses smoothing briefly as an option
considered and not adopted. If it does earn its place,
the chapter compares gridded and smoothed as parallel
rendering choices.

## Complementary example: Oregon

Oahu surfaced a specific characteristic of vector smoothing
on the first test: single-cell biome assignments become
circles. The math is correct — `ksmooth` applies Gaussian
smoothing to polygon vertices, and a four-vertex square has
its corners aggressively rounded — but the visual reads as
artifact rather than as a natural biome shape.

The artifact is most visible when isolated single cells
exist with no contiguous same-biome neighbors. Continental
regions with slow climate gradients should produce
multi-cell biome regions whose smoothing looks natural.
Oregon is the test case for this: Pacific maritime west of
the Cascades, the rain-shadowed central plateau, and the
high desert of the southeast all span enough cells at 2.5'
resolution that smoothing should produce smooth boundaries
rather than rounded blobs.

Oregon also has a strong verification surface via the
published literature — Franklin & Dyrness, *Natural
Vegetation of Oregon and Washington*, is the canonical
reference for the state's vegetation patterns.

```{r}

## Get the Oregon state polygon.
usa_states <- geodata::gadm(country = "USA",
                            level   = 1,
                            path    = "cache/gadm_cache")

oregon <- usa_states[usa_states$NAME_1 == "Oregon", ]

plot(oregon, main = "Oregon boundary from GADM")

```

Map Oregon at 2.5' resolution. The global cache is already
populated from the climate retrieval work, so no new
download is needed. Oregon is large enough (~250,000 km²)
that 2.5' (~25 km² cells) gives a respectable map without
the substantial download cost of 30 arcsec.

```{r}

oregon_map <- map_biomes(region_polygon = oregon,
                         resolution     = 2.5)

oregon_map_smooth <- smooth_biome_map(
  biome_map  = oregon_map,
  method     = "ksmooth",
  smoothness = 1
)

```

Gridded vs smoothed Oregon, each in its own chunk for full
legend visibility:

```{r fig.width=10, fig.height=6}

plot_biome_map(oregon_map_smooth, render = "grid")

```

```{r fig.width=10, fig.height=6}

plot_biome_map(oregon_map_smooth, render = "vector")

```

What to compare with the Oahu result:

- **Smoothing behavior** — Oregon's large contiguous biome
  regions (the Cascades-west forest, the eastern dry
  region) should produce smooth boundaries that read as
  natural. The single-cell artifacts visible on Oahu
  shouldn't appear on Oregon at the same smoothing
  settings, because Oregon's biome transitions span
  multiple cells.

- **Data fidelity** — the per-cell biome assignments in
  Oregon are unchanged by smoothing, same as Oahu. The
  smoothing operates only on the polygon representation.

- **Verification** — the major biome regions should match
  published vegetation maps of Oregon: temperate seasonal
  forest west of the Cascades, woodland/shrubland or
  temperate grassland east of the Cascades, with possible
  subtropical desert in the driest southeast.

The pedagogical point this comparison makes for the
chapter: smoothing is the right rendering choice for
regions whose biome transitions are continental-gradient-
sized (Oregon), and the wrong rendering choice for
regions whose interesting signal lives in small isolated
cells (Oahu's leeward deserts). The toolkit supports both;
the figure-maker chooses based on what the map needs to
show.

## Notes and limitations

- The function classifies each cell by point-in-polygon in
  T-P space using `name_biome()`. Cells whose climate falls
  outside any Whittaker biome polygon come back as NA in the
  raster.
- The function returns a list (`biome_map` class). The
  caller can plot via `plot_biome_map()` (the helper in
  `R/map_biomes.R` using terra's native plot) or build a
  ggplot from `biome_map$biome_raster` and
  `biome_map$region` directly.
- The 30 arc-second WorldClim tile download is one-time;
  subsequent calls reuse the cache at
  `cache/worldclim_cache/` (the package default). The cache
  is excluded from version control via `.gitignore`.
- Future toolkit extensions (per
  `background/chapter_mapping_ideas.md`): smoothing
  options (`smoothing = c("none", "pre", "post", "vector")`),
  boundary representation parameters parallel to
  `plot_biomes()`, KML export for Google Earth overlay.
- The next obvious worked example is Hawaii Island (the Big
  Island), where the larger canvas, stronger climate
  gradient (Hilo–Kawaihae ~14× precipitation difference),
  and elevational range to 4,200 m should produce a
  substantially richer map. Hawaii County is a single
  island administratively — no bounding-box gymnastics
  needed.