---
title: "Whittakerr — Biome Maps in 3D via Google Earth"
output: html_notebook
---

## Overview

This notebook demonstrates `export_biome_kml()` — a function
that exports a biome map as KML for viewing in Google Earth.
The KML uses vector polygons (not a raster GroundOverlay), so
the biome boundaries remain vector-quality at any zoom and
drape onto Google Earth's terrain via `altitudeMode =
clampToGround` plus `<tessellate>1</tessellate>`.

The chapter's central claim — see
`background/chapter_3d_overlay_ideas.md` Thread 1 — is that
the biome map produced from T and P alone should track
topographic features it never saw, because at scales where
orography dominates climate, both T and P track topography
(T via lapse rate, P via orographic lift and rain shadow).
Draping the 2D biome map over 3D terrain tests this prediction
by an independent physical mechanism. The classifier is
"right" in the sense that physics it doesn't know about is
visible in its output.

Oregon is the first worked example. The Cascades create a
sharp north-south barrier producing a dramatic east-west
moisture gradient — Pacific maritime west of the crest,
continental high desert east of it. The biome boundaries
derived from T and P alone should align with the Cascades
crest when draped on terrain.

## Setup

```{r}

## Load the whittakerr package. Provides export_biome_kml()
## plus the other mapping toolkit functions and bundled data.
##
## If whittakerr isn't installed yet:
##   library(devtools)
##   install_github("kimbridges/whittakerr")

library(whittakerr) ## the canonical source
library(geodata)    ## WorldClim and GADM download
library(terra)      ## raster ops, polygon ops, plot
library(sp)         ## point.in.polygon (used by name_biome)
library(sf)         ## simple features (smoothr, KML interop)
library(smoothr)    ## polygon boundary smoothing

```

## Fetching the Oregon polygon

We use GADM at level 1 (states/provinces) and filter to
Oregon. GADM caches the country-level download; first call
fetches ~3 MB.

```{r}

us_states <- geodata::gadm(country = "USA",
                           level   = 1,
                           path    = "cache/gadm_cache")

oregon <- us_states[us_states$NAME_1 == "Oregon", ]

## Verify by plotting the bare polygon before climate work.
terra::plot(oregon,
            main = "Oregon — GADM level 1 polygon",
            col  = "lightblue",
            lwd  = 0.5)

```

## Classifying biomes at 2.5 arc-minutes

We use the 2.5' global resolution. At this resolution the
WorldClim cache is already populated from earlier sessions
(Hawaii, Oregon 2D mapping). For Oregon the Cascades gradient
is well-represented at 2.5'; finer resolution (30") would
capture more local detail but isn't needed for the
orographic-verification demonstration.

```{r}

oregon_map <- map_biomes(region_polygon = oregon,
                         resolution     = 2.5)

## Quick 2D sanity check before smoothing and KML export.
plot_biome_map(oregon_map,
               render = "grid",
               legend = TRUE)

```

What to expect in the 2D map: Pacific maritime biomes
(temperate seasonal forest and temperate rain forest) west of
the Cascades crest; transitioning to woodland/shrubland on
the lee side; high desert / temperate grassland southeast.
The biome boundary follows the Cascades because the climate
inputs do.

## Smoothing for the KML export

The KML export can work directly from the raster-derived
polygons, but smoothed boundaries look more natural when
draped over terrain in Google Earth — the gridded raster
boundaries appear as stepped lines on the 3D surface.
Smoothing produces curves that read as natural biome
transitions.

For Oregon's continental-gradient regions, vector smoothing
works well (no single-cell-circle artifacts as on Oahu — see
`background/chapter_mapping_ideas.md` Thread 4 for the
discussion of when smoothing helps vs. hurts).

```{r}

oregon_map <- smooth_biome_map(oregon_map,
                               method     = "ksmooth",
                               smoothness = 1)

## Quick visual confirmation of the smoothed polygons before
## export.
plot_biome_map(oregon_map,
               render = "vector",
               legend = TRUE)

```

## Exporting to KML

Vector polygons with per-biome Style elements. The KML
includes `altitudeMode = clampToGround` and `tessellate = 1`
so Google Earth drapes each polygon over its terrain rather
than rendering it as a flat sheet at sea level.

```{r}

kml_path <- "oregon_biomes.kml"

export_biome_kml(
  biome_map   = oregon_map,
  file        = kml_path,
  smoothed    = TRUE,
  alpha       = "B0",       ## ~69% opacity
  outline     = TRUE,
  name        = "Oregon Biome Map (Whittakerr, 2.5')",
  description = "Whittaker biome classification of Oregon at 2.5 arc-minute WorldClim resolution. Polygons derived from T (BIO1) and P (BIO12) only; elevation is not an input. The orographic-verification claim predicts these boundaries should track the Cascades crest when draped on Google Earth's terrain."
)

```

The KML file is written to `oregon_biomes.kml` at the project root. To view:

1. Open Google Earth Pro (desktop) or Google Earth Web.
2. File → Open → select `oregon_biomes.kml`, or drag the file
   onto the Google Earth window.
3. Once loaded, navigate to Oregon (the polygons should auto-
   center the view). Tilt the view to see the terrain in 3D
   (mouse wheel or Shift+drag in the desktop client).

## What to look for in Google Earth

The orographic verification claim predicts specific alignments
between biome boundaries and terrain. Worth observing:

**Cascades crest.** The transition from temperate seasonal /
temperate rain forest (greens) to woodland/shrubland (drier
biomes east of the crest) should follow the Cascades ridge
line. The biome polygon edge should sit on or near the
high-elevation backbone — that's the orographic argument:
the rain shadow effect drops precipitation east of the
crest, dropping the biome assignment with it.

**Coast Range.** A secondary, weaker version of the same
effect — the Oregon Coast Range produces a small rain
shadow in the Willamette Valley. Whether the biome map
captures this depends on resolution.

**Klamath Mountains (southwest).** Topographically complex;
the biome assignment in this region tests whether the climate
raster picks up the elevational diversity.

**High desert southeast.** The Steens Mountain block and the
basin-and-range topography of southeastern Oregon. Biome
boundaries here may NOT track topography as cleanly because
the climate gradient is gentler and other factors (soil,
fire history) become relatively more important — this is
exactly the kind of region where Thread 2 of
`chapter_3d_overlay_ideas.md` (mismatches as discovery
surfaces) might surface productive ecology.

## Mismatches worth investigating

When you find biome assignments that don't match what's
actually on the ground at that location, the mismatch is
informative. The Whittaker classifier consumes T and P
only; mismatches point to factors at the next level down
the ecological hierarchy:

- **Fire-maintained ecosystems.** Eastern Oregon historically
  had extensive ponderosa pine savanna maintained by frequent
  low-intensity fire. Fire suppression has shifted many of
  these areas toward denser forest. The biome map (which
  reflects T-P climate) may predict "woodland/shrubland"
  where the actual current ecosystem is altered ponderosa
  pine.
- **Hydrological features.** Riparian gallery forests along
  the John Day, Owyhee, and other Oregon rivers exist as
  green ribbons through nominally drier biome assignments.
  The climate raster averages these out; the local ecology
  is decoupled from regional precipitation.
- **Soil chemistry.** Serpentine outcrops in the Klamath
  Mountains host distinctive endemic communities.

Each mismatch points to where ecosystem-level investigation
is warranted — fire history, hydrology, soils. The biome
map is the first cut; mismatches are the road map for the
deeper work.

## Limitations

- **Smoothing artifact at small isolated cells.** Same
  caveat as the Oahu mapping work — if Oregon has any
  single-cell biome assignments (likely a few in the
  topographically complex SW), they'll smooth to circles.
  Inspect the smoothed output before export.
- **GADM polygon simplification.** Oregon's GADM level-1
  polygon is simplified; the coastline and state boundary
  are not pixel-precise. Doesn't affect classification
  (computed from raster cells, not the polygon edge) but
  the rendered shape is slightly inside the true outline
  in places.
- **Vertical exaggeration.** Google Earth's default
  vertical exaggeration is 1.0 — true terrain. The
  Cascades are dramatic enough at 1.0 that no exaggeration
  is needed; cranking it to 2x or 3x adds visual impact
  but distorts the spatial relationships and can make the
  orographic alignment look stronger or weaker than it
  really is. Default to 1.0 for the chapter's figures.

## Toolkit extension queue

Items deferred from this initial KML implementation:

- **Per-biome description.** Each `<Placemark>` could carry a
  rich-text description visible when the user clicks the
  polygon (biome name, T-P envelope, characteristic
  vegetation, references). Currently the description is
  just the biome name.
- **Time-stamped KML.** Climate-shift scenarios via
  `get_climate(scenario = "future_ssp245")` produce future
  biome maps; KML supports `<TimeSpan>` elements that would
  let Google Earth animate the present → future shift.
- **Network links.** A master KML that references regional
  KML files would let the reader navigate to any of the
  document's worked examples from a single Google Earth
  session.
- **Raster GroundOverlay alternative.** For coarse-resolution
  global views, vector KML may be too lightweight; rasterized
  GroundOverlay would carry pixel-level color. Implement if a
  global biome map demonstration enters the document.
- **Custom altitude per biome.** KML supports per-feature
  altitude — raising forest biomes a few meters above
  clamped ground could make the 3D structure more legible,
  though it would also be slightly deceptive. Defer until
  needed.
