---
title: "Whitaker Biome Identification"
output: html_notebook
---

> **Note (2026-05-14):** The canonical source for `name_biome()`
> and `plot_biomes()` is now `R/whittakerr_functions.R`. The
> function-definition chunks below in this notebook are
> retained for documentation/teaching purposes and currently
> define the same functions inline. If you edit a function,
> edit it in `R/whittakerr_functions.R` first; the inline
> chunks here should be kept in sync (or refactored to display-
> only in a future cleanup pass — see proj_whittakerr.md
> queued tasks). `Whittakerr_climate.Rmd` sources the R/ file
> directly and does not depend on this notebook being run
> first.

## Biomes

Here is the order in the data

1, Tropical seasonal forest
2, Subtropical desert
3, Temperate rain forest
4, Tropical rain forest
5, Woodland/shrubland
6, Tundra
7, Boreal Forest
8, Temperature Grassland
9, Temperate seasonal forest

## Examples
## https://rawgit.com/valentinitnelav/plotbiomes/master/html/Whittaker_biomes_examples.html

```{r}

## Load the whittakerr package. This makes Whittaker_biomes
## (data.frame, 775 rows) and Ricklefs_colors (named character
## vector, 9 entries) available as package data, plus the
## name_biome() and plot_biomes() functions. See ?whittakerr
## for the unit convention (precipitation in cm) and the
## attribution block.
##
## If whittakerr isn't installed yet:
##   library(devtools)
##   install_github("kimbridges/whittakerr")

library(whittakerr)  ## the canonical source
library(sp)          ## point.in.polygon function
library(tidyverse)
library(ggplot2)


```

## Whittaker Plot Functions

```{r}

############################################################

name_biome <- function(mean_temp_c, total_ppt_cm){

  ## Retrieves the biome name from the temperature and
  ## precipitation. Precipitation is in centimetres (matches
  ## the underlying Whittaker_biomes$precp_cm polygon data).

  ## requires dplyr, sp

  biome_id <- NULL

  ## Extract biome names.
  id <- unique(Whittaker_biomes$biome_id)
  biome_id[id] <- unique(Whittaker_biomes$biome)

  ## Initialize the name (case: outside the diagram).
  biome_name <- "unknown"

  ## Look inside each boundary for the point.
  for(i in 1:9){
    ## Isolate one of the boundary polygons (i.e., biome).
    boundary <- Whittaker_biomes |>
      filter(Whittaker_biomes$biome_id == i)
    ## Test if the data point is inside the polygon.
    inside <- point.in.polygon(mean_temp_c,
                               total_ppt_cm,
                               boundary$temp_c,
                               boundary$precp_cm,
                               mode.checked = FALSE)

  ## If the point is inside, get the biome name.
  if(inside == 1){biome_name <- biome_id[i]}
  } ## End of boundary loop

  return(biome_name)

  } ## End function name_biome

```



```{r}

############################################################

plot_biomes <- function(mean_temp_c  = NULL,
                        total_ppt_cm = NULL,
                        source       = NULL,
                        label        = NULL){

  ## If no data points are given (e.g., mean_temp_c, total_ppt_cm),
  ##  you get just a basic plot of the Whittaker biomes.
  ##
  ## Precipitation is in centimetres (matches the underlying
  ## Whittaker_biomes$precp_cm polygon data and the diagram's
  ## native unit).

  ## Required libraries: ggplot2

  ## Create the base plot.
  W_plot <- ggplot() +
    ## Create the biome polygons.
    geom_polygon(data = Whittaker_biomes,
                 aes(x    = temp_c,
                     y    = precp_cm,
                     fill = biome),
                 # adjust polygon borders
                 colour    = "gray98",
                 linewidth = 1) +    ## was `size` pre-ggplot2 3.4.0
    ## Add axis labels.
    labs(x="Annual Mean Temperature (C)",
         y="Total Annual Precipitation (cm)") +
    ## Use a better color scheme.
    scale_fill_manual(name   = "Whittaker biomes",
                      breaks = names(Ricklefs_colors),
                      labels = names(Ricklefs_colors),
                      values = Ricklefs_colors) +
    theme_bw()

  ## Add data points if climate data are supplied.
  ##
  ## IMPORTANT: build an explicit data frame and pass it via
  ## data = ... with inherit.aes = FALSE. Without this, the
  ## geom inherits data = Whittaker_biomes (775 polygon-vertex
  ## rows) from the base ggplot call; the resulting length
  ## mismatch between the inherited data and the (usually
  ## much shorter) argument vectors silently suppresses the
  ## points.
  if(!is.null(mean_temp_c)) {
    points_df <- data.frame(mean_temp_c  = mean_temp_c,
                            total_ppt_cm = total_ppt_cm)
    W_plot <- W_plot +
      geom_point(data        = points_df,
                 aes(x = mean_temp_c, y = total_ppt_cm),
                 inherit.aes = FALSE,
                 size        = 3,
                 shape       = 21,
                 color       = "white",
                 fill        = "black",
                 stroke      = 0.7,
                 alpha       = 1.0)
    } ## End of data point addition

  ## Add label if label data are supplied.
  ## Same data-inheritance fix as the points geom above.
  if(!is.null(label)) {
    label_df <- data.frame(mean_temp_c  = mean_temp_c,
                           total_ppt_cm = total_ppt_cm,
                           label        = label)
    W_plot <- W_plot +
      geom_label(data        = label_df,
                 aes(x     = mean_temp_c,
                     y     = total_ppt_cm,
                     label = label),
                 inherit.aes = FALSE,
                 color       = "white",
                 fill        = "black",
                 alpha       = 1.0,
                 hjust       = 0,
                 nudge_x     = 1.0)
    } ## End of label addition
  
  ## Add caption at bottom if source is specified.
  if(!is.null(source)){
     W_plot <- W_plot +
       labs(caption = source)
  } ## End caption addition
  
  return(W_plot)
  
  } ## End function plot_biomes
 
```


### Testing the functions

This tests all the sorts of uses.

```{r}

## Establish the search criteria. Precipitation in cm.
mean_temp_c   <- 28.3
total_ppt_cm  <- 161

name <- name_biome(mean_temp_c = mean_temp_c, total_ppt_cm = total_ppt_cm)

paste0("The biome is: ",name,"\n")

## Plot just the biomes.
plot_biomes()

## Add a data point on the biomes.
plot_biomes(mean_temp_c = mean_temp_c, total_ppt_cm = total_ppt_cm)

## Add a data point on the biomes and a source note.
plot_biomes(mean_temp_c  = mean_temp_c,
            total_ppt_cm = total_ppt_cm,
            source       = "Source: Synthetic data")

## Plot multiple points (precipitation in cm).
t_multi <- c(5,12,14)
p_multi <- c(100,210,167)
source  <- "Testing multiple points"

plot_biomes(mean_temp_c  = t_multi,
            total_ppt_cm = p_multi,
            source       = source)

## Similar, with labels.
l_multi <- c("1","2","2a")

plot_biomes(mean_temp_c  = t_multi,
            total_ppt_cm = p_multi,
            source       = source,
            label        = l_multi)

## Use tabular data. Column ppt is in cm.
b_table <- read_csv(file =
   "site, temp, ppt
    A,     5,   100
    B,    12,   210
    C,    14,   167")

biome_plot <- plot_biomes(mean_temp_c  = b_table$temp,
                          total_ppt_cm = b_table$ppt,
                          source       = "Synthetic data table",
                          label        = b_table$site)

## Output an example diagram
ggsave(plot=biome_plot,"biome_example.png")

## Show the example diagram

biome_plot
```

