---
title: "Whittakerr — Climate Retrieval"
output: html_notebook
---

## Overview

This notebook provides `get_climate()` — a function that returns
annual mean temperature and annual precipitation at one or more
coordinate pairs, drawn from WorldClim v2.1. Both the 1970–2000
historical baseline and CMIP6 future projections are supported
under a single function signature, with defaults that make the
historical case the path of least resistance.

The output flows directly into `plot_biomes()` for placement on
the Whittaker diagram, completing the coordinate → biome pipeline.

Bioclim variables used: BIO1 (annual mean temperature, °C) and
BIO12 (annual precipitation, mm).

WorldClim is large. The first call for a given resolution and
scenario downloads the global dataset (several hundred MB at
2.5'); subsequent calls reuse the cache. The default cache
location is `data/worldclim_cache/` inside the project. Points
that fall on ocean cells return NA — handle accordingly downstream.

## Setup

```{r}

## Load the whittakerr package, which provides:
##   - Whittaker_biomes (data.frame) and Ricklefs_colors
##     (named character vector) as bundled package data
##   - name_biome(), plot_biomes(), get_climate(),
##     map_biomes(), smooth_biome_map(), plot_biome_map(),
##     and export_biome_kml() as exported functions
##
## If whittakerr isn't installed yet:
##   library(devtools)
##   install_github("kimbridges/whittakerr")

library(whittakerr) ## the canonical source
library(geodata)    ## WorldClim download (whittakerr depends on this)
library(terra)      ## raster extraction
library(tibble)     ## return type
library(dplyr)      ## piping, mutate
library(ggplot2)    ## plots package
library(sp)         ## point.in.polygon (used by name_biome internals)

```

## The Function

```{r}

############################################################

get_climate <- function(lon, lat,
                        scenario   = c("historical", "future"),
                        gcm        = "MPI-ESM1-2-HR",
                        ssp        = "245",
                        period     = "2041-2060",
                        resolution = 2.5,
                        cache_path = "data/worldclim_cache"){

  ## Returns annual mean temperature and annual precipitation
  ## at one or more (lon, lat) points, from WorldClim v2.1.
  ## Supports the 1970–2000 historical baseline (scenario =
  ## "historical", the default) and CMIP6 future projections
  ## (scenario = "future"). Future scenarios are parameterized
  ## by GCM, SSP, and time period.

  ## Returns a tibble with columns: lon, lat, mat_c (annual
  ## mean temperature, °C), map_mm (annual precipitation, mm),
  ## map_cm (annual precipitation, cm), scenario, and — when
  ## scenario = "future" — gcm, ssp, period. Ocean points and
  ## points outside the data extent return NA for mat_c, map_mm,
  ## and map_cm.

  ## requires geodata, terra, tibble, dplyr

  scenario <- match.arg(scenario)

  ## Validate inputs.
  if(length(lon) != length(lat)){
    stop("lon and lat must be the same length.")
  }

  ## Ensure cache directory exists.
  dir.create(cache_path, showWarnings = FALSE, recursive = TRUE)

  ## Fetch the appropriate bioclim SpatRaster.
  ## Historical uses var = "bio"; CMIP6 future uses var = "bioc".
  ## Layer order in both cases is BIO1 through BIO19.
  if(scenario == "historical"){
    bio_raster <- geodata::worldclim_global(
      var  = "bio",
      res  = resolution,
      path = cache_path
    )
  } else {
    bio_raster <- geodata::cmip6_world(
      model = gcm,
      ssp   = ssp,
      time  = period,
      var   = "bioc",
      res   = resolution,
      path  = cache_path
    )
  }

  ## Subset to BIO1 and BIO12.
  bio_subset <- bio_raster[[c(1, 12)]]

  ## Build a SpatVector of the query points (WGS84).
  pts <- terra::vect(cbind(lon, lat), crs = "EPSG:4326")

  ## Extract values at each point. terra::extract returns a
  ## data.frame with ID + one column per layer; values are in
  ## columns 2 and 3.
  extracted <- terra::extract(bio_subset, pts)
  mat <- extracted[, 2]
  map <- extracted[, 3]

  ## Assemble the result.
  result <- tibble(
    lon      = lon,
    lat      = lat,
    mat_c    = mat,
    map_mm   = map,
    map_cm   = map / 10,
    scenario = scenario
  )

  ## Tack on future-scenario metadata when relevant.
  if(scenario == "future"){
    result <- result |>
      mutate(gcm    = gcm,
             ssp    = ssp,
             period = period)
  }

  return(result)

} ## End function get_climate

```

## Testing — single point

A first call to confirm the function returns sensible values for
a place we know.

```{r}

## Honolulu, Hawaii.
honolulu <- get_climate(lon = -157.86, lat = 21.31)
honolulu

```

Expected: warm (mat_c around 23–25°C) and moderately wet
(map_mm around 400–800 mm depending on exactly which cell the
coordinate falls in — leeward Honolulu is drier than windward
areas a few miles away).

## Testing — vector of points

Three Pacific-coast cities of different climates, in one call.

```{r}

cities <- tibble(
  name = c("Honolulu", "Los Angeles", "Seattle"),
  lon  = c(-157.86,    -118.24,        -122.33),
  lat  = c(  21.31,      34.05,          47.61)
)

cities_historical <- get_climate(lon = cities$lon, lat = cities$lat)

## Bind the climate values back to the city names for readability.
cities_historical <- bind_cols(cities, cities_historical |>
                                 select(mat_c, map_mm, map_cm, scenario))
cities_historical

```

## Testing — future scenario

Same three cities under SSP2-4.5 (middle-of-the-road emissions),
2041–2060, MPI-ESM1-2-HR. This is the default future spec.

```{r}

cities_future <- get_climate(lon      = cities$lon,
                             lat      = cities$lat,
                             scenario = "future")

cities_future <- bind_cols(cities, cities_future |>
                             select(mat_c, map_mm, map_cm,
                                    scenario, gcm, ssp, period))
cities_future

## Side-by-side delta on temperature and precipitation.
cities_delta <- tibble(
  name           = cities$name,
  mat_historical = cities_historical$mat_c,
  mat_future     = cities_future$mat_c,
  mat_delta_c    = cities_future$mat_c   - cities_historical$mat_c,
  map_historical = cities_historical$map_mm,
  map_future     = cities_future$map_mm,
  map_delta_mm   = cities_future$map_mm  - cities_historical$map_mm
)
cities_delta

```

## Smoke test — Botanical Gardens (California)

The end-to-end pipeline: read the staged geocoded CSV, filter to
California gardens, retrieve climate, place each garden on the
Whittaker diagram. This is the first chapter-3-plus-chapter-4
integration: coordinates in, biome diagram out.

```{r}

## Load the staged dataset.
##
## fileEncoding note: the file is CP1252 / Windows-1252 encoded
## (single-byte 0xF1 for ñ in "La Cañada Flintridge" and similar).
## Reading it as UTF-8 causes scan() to throw
## "invalid input found on input connection" and return zero
## rows — silently breaking every downstream chunk. Use
## "latin1" (or equivalently "Windows-1252") to decode the
## file correctly. If the file is ever re-saved as proper
## UTF-8, this parameter should be changed back to "UTF-8".
gardens <- read.csv(
  system.file("extdata", "Bot_Garden_Geocode_CSV.csv",
              package = "whittakerr"),
  stringsAsFactors = FALSE,
  fileEncoding     = "latin1"
)

## Filter to California.
ca_gardens <- gardens |>
  filter(State == "CA")

nrow(ca_gardens)

## Retrieve climate (historical).
ca_climate <- get_climate(lon = ca_gardens$lon,
                          lat = ca_gardens$lat)

## Combine.
ca <- bind_cols(ca_gardens |> select(Name, City, State, lon, lat),
                ca_climate  |> select(mat_c, map_mm, map_cm))
ca

```

Plot the California gardens on the Whittaker diagram.
Precipitation is in centimeters throughout the public API
(matches the underlying Whittaker biome polygon data and the
diagram's native unit); pass `map_cm` from `get_climate()`'s
output, not `map_mm`. Labels are omitted intentionally — with
this many gardens, the spatial distribution of points on the
diagram carries the story.

```{r}

ca_plot <- plot_biomes(mean_temp_c  = ca$mat_c,
                       total_ppt_cm = ca$map_cm,
                       source       = "California botanical gardens · WorldClim v2.1, 1970–2000")

ca_plot

## Save the figure.
ggsave(plot     = ca_plot,
       filename = "images/california_gardens_historical.png",
       width    = 9,
       height   = 6,
       dpi      = 150)

```

## Future-scenario version of the Botanical Gardens example

Same California gardens, but with climate retrieved from the
SSP2-4.5 mid-century projection. Comparing this against the
historical plot shows which gardens are most exposed to biome
shift.

```{r}

ca_climate_future <- get_climate(lon      = ca_gardens$lon,
                                 lat      = ca_gardens$lat,
                                 scenario = "future")

ca_future <- bind_cols(ca_gardens |> select(Name, City, State, lon, lat),
                       ca_climate_future |>
                         select(mat_c, map_mm, map_cm,
                                gcm, ssp, period))
ca_future

ca_future_plot <- plot_biomes(
  mean_temp_c  = ca_future$mat_c,
  total_ppt_cm = ca_future$map_cm,
  source       = "California botanical gardens · WorldClim v2.1 CMIP6, MPI-ESM1-2-HR, SSP2-4.5, 2041–2060"
)

ca_future_plot

ggsave(plot     = ca_future_plot,
       filename = "images/california_gardens_future_ssp245_midcentury.png",
       width    = 9,
       height   = 6,
       dpi      = 150)

```

Comparison: the per-garden biome shift between historical and
future. This is the analytical promise of the document — each
garden's vulnerability is visible as movement on the diagram.

```{r}

ca_shift <- tibble(
  Name           = ca$Name,
  City           = ca$City,
  mat_historical = ca$mat_c,
  mat_future     = ca_future$mat_c,
  mat_delta_c    = ca_future$mat_c  - ca$mat_c,
  map_historical = ca$map_cm,
  map_future     = ca_future$map_cm,
  map_delta_cm   = ca_future$map_cm - ca$map_cm
)
ca_shift

```

## Notes

- The default future scenario (MPI-ESM1-2-HR, SSP2-4.5,
  2041–2060) is one defensible point estimate. A serious
  analysis would average across multiple GCMs or report
  ensemble spread; the function's parameterization supports
  that by accepting different `gcm` values in successive calls.
- WorldClim 2.5' is coarse for Hawaii's elevational gradients
  (Honolulu, Hilo, and Hawaii Volcanoes are essentially three
  different biomes within tens of kilometers). A 30" option
  would help; CHELSA at 30" might help more. Both are queued.
- The `data/worldclim_cache/` directory will grow large after
  the first runs. Add it to `.gitignore` if/when this becomes
  a packaged project.
